---
title: "Chapter 3: Calcium Data"
output: html_notebook
---

***

To run this code, make sure of having the associated txt document: 3_4_CALCIUM.txt

```{r}
data=read.table(file.choose())
```


```{r}
data
```



### Exercise 3.10

Use the calcium data in Table 3.4:

(a) Calculate S using the data matrix Y as in (3.29).
(b) Obtain R by calculating $r_{i2}$, $r_{i3}$, and $r_{23}$, as in (3.34) and (3.35).
(c) Find R using (3.37).

```{r}
Y=as.matrix(data)[,-1]

n=nrow(Y)
j=matrix(1,n,1)
ybar=t(Y)%*%j/n
J=matrix(1,n,n)
I=diag(n)
S=t(Y)%*%(I-J/n)%*%Y/(n-1)
DS=diag(sqrt(diag(S)))
R=solve(DS)%*%S%*%solve(DS)
ybar
S
R
####################
as.matrix(colMeans(Y))
cov(Y)
cor(Y)
```

### Exercise 3.11

Use the calcium data in Table 3.4:

(a) Find the generalized sample variance |S| as in (3.77). 

(b) Find the total sample variance tr(S) as in (3.78).

```{r}
det(S)
sum(diag(S))

```

### Exercise 3.17 

Define the following linear combinations for the variables in Table 3.4

$$z_{1}=y_{1}+y_{2}+y_{3}$$
$$z_{2}=2y_{1}-3y_{2}+2y_{3}$$

$$z_{3}=-y_{1}-2y_{2}-3y_{3}$$

(a) Find $z$and $S_{z}$ using (3.62) and (3.64). 

(b) Find $R_{z}$ from $S_{z}$ using (3.37).

```{r}
A=matrix(c(1,2,-1,1,-3,-2,1,2,-3),3,3)
A
zbar=A%*%ybar
SZ=A%*%S%*%t(A)
DZ=diag(sqrt(diag(SZ)))
RZ=solve(DZ)%*%SZ%*%solve(DZ)
zbar
SZ
RZ
```


### Exercise 3-11

**Part One**

3.12 Use the probe word data of Table 3.6:
(a) Find the generalized sample variance |S| as in (3.77).

```{r}

y=matrix(c(35,35,40,10,6,20,35,35,35,30,3.5,4.9,30,2.8,2.7,2.8,4.6,10.9,
           8,1.6,2.8,2.7,4.38,3.21,2.73,2.81,2.88,2.9,3.28,3.20),10)
y

S=var(y)
S
RoS=round(S,2)
RoS

dS=det(S)
dS
RodS=round(dS,2)
RodS
```

(b) Find the total sample variance tr(S) as in (3.78).


```{r}
############################## Part Two ##############################
trS=sum(diag(S))
trS
```


### Exercise 3.13 

For the probe word data in Table 3.6, find R using (3.37).


```{r}


############################## Exercise 3-13 ##############################
y=matrix(c(0.81,0.95,0.94,1.04,1,0.76,0.91,1.1,0.99,0.78,0.90,
           0.73,0.96,0.84,0.74,0.98,1.1,0.85,0.83,0.93,0.95,
           0.74,0.95,0.97,0.72,1.11,1.2,1.13,1,0.78,1,1,0.71,0.76,0.89,0.88,1.17,
           0.85,0.97,1,1,0.89,0.98,0.78,0.74,0.91,80,97,105,90,90,86,100,85,
           97,97,91,87,78,90,86,80,90,99,85,90,90,88,95,90,92,74,98,100,86,98,
           70,99,75,90,85,99,100,78,106,98,102,90,94,80,93,86,
           356,289,319,356,323,381,350,301,379,296,353,306,290,371,
           312,393,364,359,296,345,378,304,347,327,386,365,365,352,325,321,360,
           336,352,353,373,376,367,335,396,277,378,360,291,269,318,328,124,117,
           143,199,240,157,221,186,142,131,221,178,136,200,208,202,152,185,116,
           123,136,134,184,192,279,228,145,172,179,222,134,143,169,263,174,134,
           182,241,128,222,165,282,94,121,73,106,55,76,105,108,143,165,119,105,
           98,94,53,66,142,93,68,102,76,37,60,50,47,50,91,124,74,235,158,140,
           145,99,90,105,32,165,78,80,54,175,80,186,117,160,71,29,42,56),46)
y

S=var(y)
S
RoS=round(S,3)
RoS

D=diag(diag(RoS))
D

DI=solve(D)
DI

R=DI%*%S%*%DI
R
RoR=round(R,3)
RoR

```

